Best Practices for Alchemical Free Energy Calculations [Article v\versionnumber]

### 2021-03-17 paper/manuscript.tex

After an alchemical calculation is performed, which generally involves multiple simulations at a variety of alchemical states, the data must be analyzed to compute an estimate of the free energy for the transformation of interest.
 Early work used simple but statistically suboptimal estimators for this: free energy perturbation (FEP) used a simple (but highly biased) estimator based on the Zwanzig relation~\cite{zwanzig1954hightemperature} or numerical quadrature via thermodynamic integration (TI), for which the theory dates back the better part of a century but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamicsa, merz1989free}. %
 More recent developments have seen new, highly efficient statistical estimators that make better use of all the data, often building on the more efficient and less biased Bennett acceptance ratio (BAR)~\cite{bennett1976efficient}, producing multistate generalizations~\cite{shirts2008statisticallya} or removing the need for global equilibrium~\cite{wu2016multiensemble, mey2014xtram, wu2014statistically}.

@@

\paragraph{Simulating binding events is computationally expensive}
While simulating binding events has been used to estimate binding affinities~\cite{jong2011determining,pan2017quantitative} or to get insights into the binding pathways and kinetics of receptor-ligand systems~\cite{teo2016adaptive,votapka2017seekr,doerr2014onthefly,plattner2015protein,dixon2018predicting}, the computational cost of these calculations is usually dominated by the rate of dissociation, which can be on the microsecond timescale even for millimolar binders~\cite{pan2017quantitative} and reaches the microsecond to second timescale for a typical drug~\cite{basavapathruni2012conformational,hyre2006cooperative}.
Depending on system size and simulation settings, common molecular dynamics software packages can reach a few hundreds of ns/day using currently available high-end GPUs~\cite{eastman2017openmm,kutzner2019more}, making these type of calculations unappealing and irrelevant on a pharmaceutical drug discovery timescale.
Other methods compute the free energy of binding by building potential of mean force profiles along a reaction coordinate~\cite{woo2005calculationa,velez-vega2013overcoming,limongelli2013funnel,heinzelmann2017attachpullrelease}, but these methods require prior knowledge of a high-probability binding pathway, which is not easily available, especially in the prospective scenarios typical of the drug development process.

@@

\paragraph{Why do alchemical calculations need unphysical intermediate states?}
While it is theoretically possible to estimate the ratio of partition functions from samples collected only at states $\vec{\lambda}_A$ and $\vec{\lambda}_B$, the efficiency of the free energy estimators rapidly decreases as the phase-space overlap between the two states also decreases~\cite{wu2005phasespaceb,wu2005phasespacec}.
Roughly, the phase-space overlap between two thermodynamic states measures the degree to which high-probability configurations (i.e., those with very negative potential energy) in one state are also high-probability configurations in the other state (see Sec.~\ref{sec:are-they-good} and Fig.~\ref{fig:fig_what_is_lambda}).

Equilibrium free energy calculations, our focus here, solve the problem of having poor overlap between the states of interest by introducing multiple intermediate alchemical states at values $\vec{\lambda}_A = \vec{\lambda}_0, \vec{\lambda}_1, \cdots, \vec{\lambda}_K = \vec{\lambda}_B$ so that each pair of consecutive states $\vec{\lambda}_k, \vec{\lambda}_{k+1}$ share good overlap.
Each intermediate state models a ligand that is neither $A$ nor $B$ but a interpolation of the two.
Many estimators (e.g., exponential reweighting (EXP)~\cite{zwanzig1954hightemperature} and Bennett's acceptance ratio (BAR)~\cite{bennett1976efficient,shirts2003equilibriuma}) can then be used to compute the free energy as
\begin{equation}
    \Delta G_{\mathrm{env}} = k_BT \sum_{k=0}^{K-1} \Delta f(\vec{\lambda}_k, \vec{\lambda}_{k+1}),
\end{equation}
from samples collected at all the alchemical states $\{\vec{\lambda}_k \}$, where $\Delta f$ is the \emph{unitless free energy difference}
\begin{equation}
    \Delta f(\vec{\lambda}_k, \vec{\lambda}_{k+1}) = f(\vec{\lambda}_{k+1}) - f(\vec{\lambda}_k) = - \ln \frac{Z(\vec{\lambda}_{k+1})}{Z(\vec{\lambda}_k)} \, .
\end{equation}
While this strategy usually results in sampling thermodynamic states whose Boltzmann distributions are very similar, thus collecting information that is to some degree redundant, some estimators, such as the Multistate Bennett acceptance ratio (MBAR)~\cite{shirts2008statisticallya}, can exploit similarities between states to improve the precision of the estimates. This is achieved by using the configurations sampled at all alchemical states $\{\vec{\lambda}_k \}$ to compute the free energy difference $\Delta f(\vec{\lambda}_i, \vec{\lambda}_{j})$ between any pair of states $i,j$ (see Sec.~\ref{subsec:estimators}).

@@

Here we provide a simple estimate of the value provided by alchemical
free energy calculations in lead optimization. Let $P(\Delta\Delta
G)$ be the probability distribution of the changes in the binding free
energies of a new set of molecules during one round of lead
optimization, and let $P(\Delta\Delta G^\dagger|\Delta\Delta G)$ be the
conditional probability of the binding free energy change computed by
the free energy calculations, $\Delta\Delta G^\dagger$, given the actual
change $\Delta\Delta G$. The latter conditional probability can be modeled
by a normal distribution
\begin{equation}
  P(\Delta\Delta G^\dagger|\Delta\Delta G) = \frac{1}{\sqrt{2\pi\sigma^2}}
  \exp\left(-\frac{(\Delta\Delta G^\dagger - \Delta\Delta G)^2}{2\sigma^2}\right),
  \label{eqn:free-energy-distribution}
\end{equation}
where $\sigma$ signifies the accuracy of free energy calculations.
Here we assume that there is no systematic bias in the free energy
calculations, i.e., on average, the free energy change computed by
free energy calculations agrees with the actual free energy change. Additional analysis of this type is presented in Brown et al.~\cite{shirts2010}

@@

It is crucial when proposing compounds for synthesis to have some idea of the underlying error or uncertainty in the predictions. A retrospective assessment can give an indication of prospective performance for similar molecules~\cite{ciordia2016application}. Beyond this, several parameters provide useful indicators of performance. For example error estimates provided by free energy estimators that are too large can highlight poorly converged simulations~\cite{perez-benito2019predicting}. Hysteresis, either within cycles in the perturbation network or between forward and backward perturbations can be checked~\cite{wang2013modeling} to indicate problematic perturbations involved in cycles connecting many compounds (See also Secs.~\ref{sec:relative-fe-protocol} and~\ref{sec:are-they-good}). Once synthesis and testing of compounds is complete a standard strategy is to look back at how the calculations performed. In this regard it is important to consider the issue of selection bias upfront. It is tempting to only synthesize the compounds predicted to be most active, thus a narrow range of calculated activity is tested that imposes limits on the statistical assessment of performance, ideally example molecules from across the range of predicted activity can be assessed or corrections can be applied based on previous recommendations~\cite{abel2017critical}. For a more detailed discussion on checking the robustness of your alchemical free energy calculation see also Sec.~\ref{sec:are-they-good}.

@@

In contrast, in a dual topology alchemical free energy calculation, \emph{no atoms are allowed to change type}~\cite{boresch1999rolea, shirts2012best}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxymethyl group, then passing through an intermediate state where some atoms are partially interacting---particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}. The transformation finally culminates in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:fig_topology}'s right branch depicts how such a dual topology works.

@@

\paragraph{Constraints and relative free energy calculations}
One issue which requires particular care is the use of constraints.
Commonly, bonds involving hydrogen are constrained to a fixed length using algorithms such as SHAKE or LINCS, allowing the use of longer timesteps~\cite{krautler2001fast}.
However, in single topology relative free energy calculations, the atoms involved might be mutated to other atom types---for example, in a mutation of methane to methanol, one hydrogen might become an oxygen atom. The bonds with such atoms might not have any constraints, or if all bond are constrained, would have constraints of different lengths. 
Some molecular dynamics engines are not set up to recognize this change, or at least not to correctly include contributions to the free energy from changing constraints/constraint length, so results for a transformation can be erroneous.
At present the most general solution to this problem is simply to avoid the use of constraints (and thus use a smaller timestep if necessary, usually of around 1 fs) in any relative free energy calculation involving a transformation of a constrained bond. Individual software programs and settings can handle such issues. For example, bonds can be transformed in both GROMACS and GROMOS, because contributions of LINCS (GROMACS) and SHAKE (GROMOS) constrained bonds are added to the $\frac{dH}{d\lambda}$ term~\cite{pearlman1993determining, straatsma1992holonomica, pearlman1991overlooked, gunsteren1989computer}. However, the constraints are not taken into account when calculating energy differences to other intermediate states as the effect is entropic, not energetic. User manuals should carefully checked for how these effects are included if a constrained bond changes in length. We do note that if one performs calculations of the changing constraint in both solution and in the binding site, then the errors often cancel substantially, but the cancellation is not guaranteed.

@@

Alternatively, a set of restraints proposed by Boresch have also commonly been employed, where all six rigid-body degrees of freedom governing the orientation of the ligand relative to the receptor are restrained~\cite{boresch2003absolutea, leitgeb2005alchemicala}.
Further restraints, such as on the overall ligand RMSD have also been used~\cite{woo2005calculationa}.

@@

In absolute calculations, this restraint can substitute or be added to the restraint used to handle the standard state correction (Sec.~\ref{sec:standardstate-restraints}).
In the latter case, however, care must be taken when computing the standard state correction.
When multiple restraining potentials are active in the non-interacting state, the correction can generally be computed only through numerical integration.
Alternatively, one can adopt a protocol that removes the bound-state restraint in the non-interacting state.
Finally, even for tight binders, dissociation events can be enhanced by methods such as Hamiltonian replica exchange~\cite{sugita2000multidimensionala,chodera2011replica,wang2013identifying} and expanded ensemble~\cite{lyubartsev1992newa,li2007simulated}, especially in absolute free energy calculations using harmonic or flat-bottom restraints.
In the latter case, dissociations can be averted simply by increasing the spring constant and/or reducing the radius of the restraint potential to prevent the exploration of ligand conformations outside the binding site in the decoupled state (bottom-right state in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}) that could be propagated to the bound state.

@@

Other issues, such as whether absolute calculations should retain or remove intramolecular non-bonded interactions
through either annihilation~\cite{hermans1997inclusiona, mann2000modelinga, boresch2003absolutea, wang2006absolutea, mobley2006use} or decoupling~\cite{fujitani2005directa, mobley2006use}, must be considered. Reasonable efficiency can be often obtained with either choice even if some are somewhat better or worse than others, and there is no consensus on which is better in most given situations. Our recommendation is to leave the intramolecular interactions on during the transformation for simplicity if there are no other known issues with this approach. The key feature of the simulation to watch out for is whether the total potential energy, and therefore the intermediate ensembles sampled, changes smoothly from beginning to end. Problems of discontinuous changes of the potential energy can be diagnosed by noticing lack of configuration space overlap between different simulations (see Sec.~\ref{sec:are-they-good}).

@@

A key additional consideration in choosing the alchemical pathway is the choice of spacing of intermediate states.
The spacing depends to some extent on the choice of analysis method, though states should essentially be spaced equidistant in the relevant thermodynamic length~\cite{crooks2007measuringa, sivak2012thermodynamic}.
For BAR/MBAR techniques this means that states should be spaced so that the statistical uncertainties between neighboring states be approximately equal~\cite{pham2012optimal, shenfeld2009minimizing}, where "approximately" is roughly within 30-50\% in magnitude. 
Some schemes to adaptively optimize the spacing of intermediate states based on initial exploratory simulations have been proposed~\cite{hayes2017adaptive}. For molecules changing in dense solvent, then the best path is roughly independent of molecule size and shape, so what works for one molecular transformation is likely to be relatively efficient for another~\cite{monroe2014converging}.

Some approaches have attempted to find alternative pathways to improve efficiency or find paths of low thermodynamic length~\cite{naden2014linear,naden2015linear,pham2012optimal}. For example, the enveloping distribution sampling (EDS) approach, and its multiple-replica and accelerated variant, works to improve efficiency by creating a single artificial intermediate state which simultaneously samples all end state phase spaces~\cite{perthold2018accelerateda,sidler2017efficientb, christ2007envelopinga}. When this can be done, it provides an extremely efficient way to calculate the relative free energy difference between multiple ligands from a single simulation. However, it can often fail whenever the simulation of this intermediate state ends up trapped in configurations characteristic of only one end state. Thus, successful use of EDS can require system-dependent tuning, making it difficult to implement in an automated and reliable way. However, when successful, it can be very efficient.

@@

\subsubsection{Which sampling scheme will work best for my problem?}
\label{sec:sampling_schemes}
Though all alchemical simulations must sample from multiple $\vec{\lambda}$ states, different approaches can be used to achieve this. Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes. The simplest approach involves running an independent simulation at each of the predefined $\vec{\lambda}$ values (see Fig.~\ref{fig:fig_sampling_scheme}\textbf{A}). This type of scheme is currently used for AMBER TI calculations~\cite{song2019using} and for Sire as implemented in BioSimSpace~\cite{hedges2019biosimspace}. However, if these simulations can be run simultaneously with communication between them, a simple extension allows mixing between these replicas. In this approach, the simulation at each $\vec{\lambda}$ can undergo periodic exchanges with neighboring $\vec{\lambda}$ values. This form of replica exchange, called Hamiltonian replica exchange, is based on ideas developed from Monte Carlo simulations of spin glasses by Swendsen and Wang~\cite{swendsen1986replica}. With the Metropolis-Hastings acceptance criterion for exchanges, the generated ensemble of all replicas still samples from the Boltzmann distribution for each replica. This approach has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensionala,sugita1999replicaexchangea, woods2003developmenta, jiang2010free}. The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme}\textbf{B}. It is supported in various software packages that provide alchemical implementations, such as GROMACS~\cite{aldeghi2015accurate}, GROMOS~\cite{hritz2008hamiltoniana,hritz2007optimizationa}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}.

A third approach borrows ideas from simulated tempering~\cite{marinari1992simulateda}. In this scheme a single replica rapidly explores all of $\vec{\lambda}$ space by working out optimal weights that allow switching between different intermediate $\vec{\lambda}$ values, as seen in Fig.~\ref{fig:fig_sampling_scheme}\textbf{C} . This approach is also referred to as self-adjusted mixture sampling~\cite{lyubartsev1992newa, li2007simulated, tan2017optimally} and while promising, has so far only been supported in OpenMM Tools~\cite{andrearizzi2019choderalab} and GROMACS.  Although this approach allows multiple states to be simulated in a single simulation, the weights do not always converge to the proper equilibrium distribution, and care must be taken that the final results are converged.

@@

The (reduced) free energy difference $\Delta f_{01}$ between an initial state 0 and a final state 1 defined by two different potential energy functions 
$u_0(\vec{q})$ and $u_1(\vec{q})$ over coordinate space $\vec{q}$ can be calculated as:
\begin{eqnarray}
\Delta f_{01} & = & -\ln \expect{e^{-(u_1(\vec{q}) - u_0(\vec{q}))}}_0 =  -\ln \expect{e^{-\Delta u(\vec{q})} }_0
\end{eqnarray}\label{eqn.zwanzig}
and the average is over all samples from the simulation performed with $u_0$. In the case of NVT (canonical) sampling and assuming the masses do not change, then $u$ is simply $U/k_BT$, and $f$ is $F/k_BT$, but it can be generalized to other ensembles with the proper definition of $f$ and $u$.
Described in words, we take the samples generated during our run with the potential energy function $u_0(\vec{q})$ and calculate what the difference in energy would be if we switched instantaneously to the potential energy function $u_1(\vec{q})$, and average the exponential of the negative energy difference to get the negative of the exponential of the free energy difference. The original distributions, P($u_0$) as generated at $\vec{\lambda}=0$ and P($u_1$) would look like those seen in Fig.~\ref{fig:fig_sampling_scheme}\textbf{A}-\textbf{C} on the right hand side. Reevaluating requires almost no extra code functionality to perform; one need only to save a full precision trajectory, and run an unmodified molecular simulation code using the $u_1$ in order to calculate the new energies of stored snapshots. The analysis can be written in a line of code. We note that this method is even more general, in that the instantaneous work to change the potential energy function from $u_0$ to $u_1$ can be replaced by the non-reversible work $W$ to make the same change beginning from the same equilibrium conditions at either end state~\cite{jarzynski1997nonequilibrium,jarzynski1998equilibrium,crooks2000pathensemble}, allowing for non-equilibrium free energy calculations, an alternate approach. We do not detail non-equilibrium transformations here, and refer the reader to more advanced treatments~\cite{maragakis2008bayesian,oberhofer2005biased,procacci2015unbiased,shirts2003equilibriuma,ytreberg2004singleensemble, gapsys2020large}, as our focus here is on equilibrium free energy techniques.

Although the Zwanzig equation is formally correct as long as the two states considered sample the same phase space volume, which is true for standard molecular models, it has some very important numerical issues that mean that it often performs badly for standard free energy calculations, even for small molecules~\cite{shirts2005comparison,lu2003appropriate}. One can show that if the standard deviation of the difference $\Delta u(\vec{q}) = u_1(\vec{q})-u_2(\vec{q})$ over all sampled $\vec{q}$ is large, which in this case, means only several times $k_BT$, then very few samples contribute to the average, and the answer will be both biased and extremely noisy~\cite{lelievre2010free}. Essentially, the method is dominated by contributions of rare snapshots~\cite{jarzynski2006rare, wu2005phasespaceb, wu2005phasespacec}.

>>>>

After an alchemical calculation is performed, which generally involves multiple simulations at a variety of alchemical states, the data must be analyzed to compute an estimate of the free energy for the transformation of interest.
 Early work used simple but statistically suboptimal estimators for this: free energy perturbation (FEP) used a simple (but highly biased) estimator based on the Zwanzig relation~\cite{zwanzig1954hightemperature} or numerical quadrature via thermodynamic integration (TI), for which the theory dates back the better part of a century but with the first computational applications emerging in the 1980's and 90's~\cite{kirkwood1935statistical, jorgensen1985monte, kollman1993free, wong1986dynamics, merz1989free}. %
 More recent developments have seen new, highly efficient statistical estimators that make better use of all the data, often building on the more efficient and less biased Bennett acceptance ratio (BAR)~\cite{bennett1976efficient}, producing multistate generalizations~\cite{shirts2008statisticallya} or removing the need for global equilibrium~\cite{wu2016multiensemble, mey2014xtram, wu2014statistically}.

@@

\paragraph{Simulating binding events is computationally expensive}
While simulating binding events has been used to estimate binding affinities~\cite{jong2011determining,pan2017quantitative} or to get insights into the binding pathways and kinetics of receptor-ligand systems~\cite{teo2016adaptive,votapka2017seekr,doerr2014onthefly,plattner2015protein,dixon2018predicting}, the computational cost of these calculations is usually dominated by the rate of dissociation, which can be on the microsecond timescale even for millimolar binders~\cite{pan2017quantitative} and reaches the microsecond to second timescale for a typical drug~\cite{basavapathruni2012conformational,hyre2006cooperative}.
Depending on system size and simulation settings, common molecular dynamics software packages can reach a few hundreds of ns/day using currently available high-end GPUs~\cite{eastman2017openmm,kutzner2019more}, making these type of calculations unappealing and irrelevant on a pharmaceutical drug discovery timescale.
Other methods compute the free energy of binding by building potential of mean force profiles along a reaction coordinate~\cite{woo2005calculation,velez-vega2013overcoming,limongelli2013funnel,heinzelmann2017attachpullrelease}, but these methods require prior knowledge of a high-probability binding pathway, which is not easily available, especially in the prospective scenarios typical of the drug development process.

@@

\paragraph{Why do alchemical calculations need unphysical intermediate states?}
While it is theoretically possible to estimate the ratio of partition functions from samples collected only at states $\vec{\lambda}_A$ and $\vec{\lambda}_B$, the efficiency of the free energy estimators rapidly decreases as the phase-space overlap between the two states also decreases~\cite{wu2005phasespace, wu2005phasespacea}.
Roughly, the phase-space overlap between two thermodynamic states measures the degree to which high-probability configurations (i.e., those with very negative potential energy) in one state are also high-probability configurations in the other state (see Sec.~\ref{sec:are-they-good} and Fig.~\ref{fig:fig_what_is_lambda}).

Equilibrium free energy calculations, our focus here, solve the problem of having poor overlap between the states of interest by introducing multiple intermediate alchemical states at values $\vec{\lambda}_A = \vec{\lambda}_0, \vec{\lambda}_1, \cdots, \vec{\lambda}_K = \vec{\lambda}_B$ so that each pair of consecutive states $\vec{\lambda}_k, \vec{\lambda}_{k+1}$ share good overlap.
Each intermediate state models a ligand that is neither $A$ nor $B$ but a interpolation of the two.
Many estimators (e.g., exponential reweighting (EXP)~\cite{zwanzig1954hightemperature} and Bennett's acceptance ratio (BAR)~\cite{bennett1976efficient,shirts2003equilibrium}) can then be used to compute the free energy as
\begin{equation}
    \Delta G_{\mathrm{env}} = k_BT \sum_{k=0}^{K-1} \Delta f(\vec{\lambda}_k, \vec{\lambda}_{k+1}),
\end{equation}
from samples collected at all the alchemical states $\{\vec{\lambda}_k \}$, where $\Delta f$ is the \emph{unitless free energy difference}
\begin{equation}
    \Delta f(\vec{\lambda}_k, \vec{\lambda}_{k+1}) = f(\vec{\lambda}_{k+1}) - f(\vec{\lambda}_k) = - \ln \frac{Z(\vec{\lambda}_{k+1})}{Z(\vec{\lambda}_k)} \, .
\end{equation}
While this strategy usually results in sampling thermodynamic states whose Boltzmann distributions are very similar, thus collecting information that is to some degree redundant, some estimators, such as the Multistate Bennett acceptance ratio (MBAR)~\cite{shirts2008statisticallya}, can exploit similarities between states to improve the precision of the estimates. This is achieved by using the configurations sampled at all alchemical states $\{\vec{\lambda}_k \}$ to compute the free energy difference $\Delta f(\vec{\lambda}_i, \vec{\lambda}_{j})$ between any pair of states $i,j$ (see Sec.~\ref{subsec:estimators}).

@@

Here we provide a simple estimate of the value provided by alchemical
free energy calculations in lead optimization. Let $P(\Delta\Delta
G)$ be the probability distribution of the changes in the binding free
energies of a new set of molecules during one round of lead
optimization, and let $P(\Delta\Delta G^\dagger|\Delta\Delta G)$ be the
conditional probability of the binding free energy change computed by
the free energy calculations, $\Delta\Delta G^\dagger$, given the actual
change $\Delta\Delta G$. The latter conditional probability can be modeled
by a normal distribution
\begin{equation}
  P(\Delta\Delta G^\dagger|\Delta\Delta G) = \frac{1}{\sqrt{2\pi\sigma^2}}
  \exp\left(-\frac{(\Delta\Delta G^\dagger - \Delta\Delta G)^2}{2\sigma^2}\right),
  \label{eqn:free-energy-distribution}
\end{equation}
where $\sigma$ signifies the accuracy of free energy calculations.
Here we assume that there is no systematic bias in the free energy
calculations, i.e., on average, the free energy change computed by
free energy calculations agrees with the actual free energy change. Additional analysis of this type is presented in Brown et al.~\cite{brown2010freeenergy}

@@

It is crucial when proposing compounds for synthesis to have some idea of the underlying error or uncertainty in the predictions. A retrospective assessment can give an indication of prospective performance for similar molecules~\cite{ciordia2016application}. Beyond this, several parameters provide useful indicators of performance. For example error estimates provided by free energy estimators that are too large can highlight poorly converged simulations~\cite{perez-benito2019predicting}. Hysteresis, either within cycles in the perturbation network or between forward and backward perturbations can be checked~\cite{wang2013modeling} to indicate problematic perturbations involved in cycles connecting many compounds (See also Secs.~\ref{sec:relative-fe-protocol} and~\ref{sec:are-they-good}). Once synthesis and testing of compounds is complete a standard strategy is to look back at how the calculations performed. In this regard it is important to consider the issue of selection bias upfront. It is tempting to only synthesize the compounds predicted to be most active, thus a narrow range of calculated activity is tested that imposes limits on the statistical assessment of performance, ideally example molecules from across the range of predicted activity can be assessed or corrections can be applied based on previous recommendations~\cite{robert2017critical}. For a more detailed discussion on checking the robustness of your alchemical free energy calculation see also Sec.~\ref{sec:are-they-good}.

@@

In contrast, in a dual topology alchemical free energy calculation, \emph{no atoms are allowed to change type}~\cite{boresch1999role, shirts2012best}. This means that the benzene to benzyl alcohol transformation involves starting with benzene plus the non-interacting dummy atoms making up the hydroxymethyl group, then passing through an intermediate state where some atoms are partially interacting---particularly, those atoms which are becoming dummy atoms or ceasing to be dummy atoms~\cite{mobley2014blind}. The transformation finally culminates in a state where benzyl alcohol is present along with the additional dummy atom which was previously a corresponding hydrogen of the benzene. Fig.~\ref{fig:fig_topology}'s right branch depicts how such a dual topology works.

@@

\paragraph{Constraints and relative free energy calculations}
One issue which requires particular care is the use of constraints.
Commonly, bonds involving hydrogen are constrained to a fixed length using algorithms such as SHAKE or LINCS, allowing the use of longer timesteps~\cite{krautler2001fast}.
However, in single topology relative free energy calculations, the atoms involved might be mutated to other atom types---for example, in a mutation of methane to methanol, one hydrogen might become an oxygen atom. The bonds with such atoms might not have any constraints, or if all bond are constrained, would have constraints of different lengths. 
Some molecular dynamics engines are not set up to recognize this change, or at least not to correctly include contributions to the free energy from changing constraints/constraint length, so results for a transformation can be erroneous.
At present the most general solution to this problem is simply to avoid the use of constraints (and thus use a smaller timestep if necessary, usually of around 1 fs) in any relative free energy calculation involving a transformation of a constrained bond. Individual software programs and settings can handle such issues. For example, bonds can be transformed in both GROMACS and GROMOS, because contributions of LINCS (GROMACS) and SHAKE (GROMOS) constrained bonds are added to the $\frac{dH}{d\lambda}$ term~\cite{pearlman1993determining, straatsma1992holonomic, pearlman1991overlooked, gunsteren1989computer}. However, the constraints are not taken into account when calculating energy differences to other intermediate states as the effect is entropic, not energetic. User manuals should carefully checked for how these effects are included if a constrained bond changes in length. We do note that if one performs calculations of the changing constraint in both solution and in the binding site, then the errors often cancel substantially, but the cancellation is not guaranteed.

@@

Alternatively, a set of restraints proposed by Boresch have also commonly been employed, where all six rigid-body degrees of freedom governing the orientation of the ligand relative to the receptor are restrained~\cite{boresch2003absolute, leitgeb2005alchemical}.
Further restraints, such as on the overall ligand RMSD have also been used~\cite{woo2005calculation}.

@@

In absolute calculations, this restraint can substitute or be added to the restraint used to handle the standard state correction (Sec.~\ref{sec:standardstate-restraints}).
In the latter case, however, care must be taken when computing the standard state correction.
When multiple restraining potentials are active in the non-interacting state, the correction can generally be computed only through numerical integration.
Alternatively, one can adopt a protocol that removes the bound-state restraint in the non-interacting state.
Finally, even for tight binders, dissociation events can be enhanced by methods such as Hamiltonian replica exchange~\cite{sugita2000multidimensional,chodera2011replica,wang2013identifying} and expanded ensemble~\cite{lyubartsev1992new,li2007simulated}, especially in absolute free energy calculations using harmonic or flat-bottom restraints.
In the latter case, dissociations can be averted simply by increasing the spring constant and/or reducing the radius of the restraint potential to prevent the exploration of ligand conformations outside the binding site in the decoupled state (bottom-right state in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle}) that could be propagated to the bound state.

@@

Other issues, such as whether absolute calculations should retain or remove intramolecular non-bonded interactions
through either annihilation~\cite{hermans1997inclusion, mann2000modeling, boresch2003absolute, wang2006absolute, mobley2006use} or decoupling~\cite{fujitani2005direct, mobley2006use}, must be considered. Reasonable efficiency can be often obtained with either choice even if some are somewhat better or worse than others, and there is no consensus on which is better in most given situations. Our recommendation is to leave the intramolecular interactions on during the transformation for simplicity if there are no other known issues with this approach. The key feature of the simulation to watch out for is whether the total potential energy, and therefore the intermediate ensembles sampled, changes smoothly from beginning to end. Problems of discontinuous changes of the potential energy can be diagnosed by noticing lack of configuration space overlap between different simulations (see Sec.~\ref{sec:are-they-good}).

@@

A key additional consideration in choosing the alchemical pathway is the choice of spacing of intermediate states.
The spacing depends to some extent on the choice of analysis method, though states should essentially be spaced equidistant in the relevant thermodynamic length~\cite{crooks2007measuring, sivak2012thermodynamic}.
For BAR/MBAR techniques this means that states should be spaced so that the statistical uncertainties between neighboring states be approximately equal~\cite{pham2012optimal, shenfeld2009minimizing}, where "approximately" is roughly within 30-50\% in magnitude. 
Some schemes to adaptively optimize the spacing of intermediate states based on initial exploratory simulations have been proposed~\cite{hayes2017adaptive}. For molecules changing in dense solvent, then the best path is roughly independent of molecule size and shape, so what works for one molecular transformation is likely to be relatively efficient for another~\cite{monroe2014converging}.

Some approaches have attempted to find alternative pathways to improve efficiency or find paths of low thermodynamic length~\cite{naden2014linear,naden2015linear,pham2012optimal}. For example, the enveloping distribution sampling (EDS) approach, and its multiple-replica and accelerated variant, works to improve efficiency by creating a single artificial intermediate state which simultaneously samples all end state phase spaces~\cite{perthold2018accelerated,sidler2017efficient, christ2007enveloping}. When this can be done, it provides an extremely efficient way to calculate the relative free energy difference between multiple ligands from a single simulation. However, it can often fail whenever the simulation of this intermediate state ends up trapped in configurations characteristic of only one end state. Thus, successful use of EDS can require system-dependent tuning, making it difficult to implement in an automated and reliable way. However, when successful, it can be very efficient.

@@

\subsubsection{Which sampling scheme will work best for my problem?}
\label{sec:sampling_schemes}
Though all alchemical simulations must sample from multiple $\vec{\lambda}$ states, different approaches can be used to achieve this. Fig.~\ref{fig:fig_sampling_scheme} illustrates the four most common schemes. The simplest approach involves running an independent simulation at each of the predefined $\vec{\lambda}$ values (see Fig.~\ref{fig:fig_sampling_scheme}\textbf{A}). This type of scheme is currently used for AMBER TI calculations~\cite{song2019using} and for Sire as implemented in BioSimSpace~\cite{hedges2019biosimspace}. However, if these simulations can be run simultaneously with communication between them, a simple extension allows mixing between these replicas. In this approach, the simulation at each $\vec{\lambda}$ can undergo periodic exchanges with neighboring $\vec{\lambda}$ values. This form of replica exchange, called Hamiltonian replica exchange, is based on ideas developed from Monte Carlo simulations of spin glasses by Swendsen and Wang~\cite{swendsen1986replica}. With the Metropolis-Hastings acceptance criterion for exchanges, the generated ensemble of all replicas still samples from the Boltzmann distribution for each replica. This approach has been used in many different contexts for molecular simulations~\cite{sugita2000multidimensional,sugita1999replicaexchange, woods2003development, jiang2010free}. The basic idea of the replica exchange scheme is shown in Fig.~\ref{fig:fig_sampling_scheme}\textbf{B}. It is supported in various software packages that provide alchemical implementations, such as GROMACS~\cite{aldeghi2015accurate}, GROMOS~\cite{hritz2008hamiltonian,hritz2007optimization}, FEP+~\cite{wang2015accurate}, and NAMD~\cite{jiang2019computing}.

A third approach borrows ideas from simulated tempering~\cite{marinari1992simulated}. In this scheme a single replica rapidly explores all of $\vec{\lambda}$ space by working out optimal weights that allow switching between different intermediate $\vec{\lambda}$ values, as seen in Fig.~\ref{fig:fig_sampling_scheme}\textbf{C} . This approach is also referred to as self-adjusted mixture sampling~\cite{lyubartsev1992new, li2007simulated, tan2017optimally} and while promising, has so far only been supported in OpenMM Tools~\cite{andrearizzi2019choderalab} and GROMACS.  Although this approach allows multiple states to be simulated in a single simulation, the weights do not always converge to the proper equilibrium distribution, and care must be taken that the final results are converged.

@@

The (reduced) free energy difference $\Delta f_{01}$ between an initial state 0 and a final state 1 defined by two different potential energy functions 
$u_0(\vec{q})$ and $u_1(\vec{q})$ over coordinate space $\vec{q}$ can be calculated as:
\begin{eqnarray}
\Delta f_{01} & = & -\ln \expect{e^{-(u_1(\vec{q}) - u_0(\vec{q}))}}_0 =  -\ln \expect{e^{-\Delta u(\vec{q})} }_0
\end{eqnarray}\label{eqn.zwanzig}
and the average is over all samples from the simulation performed with $u_0$. In the case of NVT (canonical) sampling and assuming the masses do not change, then $u$ is simply $U/k_BT$, and $f$ is $F/k_BT$, but it can be generalized to other ensembles with the proper definition of $f$ and $u$.
Described in words, we take the samples generated during our run with the potential energy function $u_0(\vec{q})$ and calculate what the difference in energy would be if we switched instantaneously to the potential energy function $u_1(\vec{q})$, and average the exponential of the negative energy difference to get the negative of the exponential of the free energy difference. The original distributions, P($u_0$) as generated at $\vec{\lambda}=0$ and P($u_1$) would look like those seen in Fig.~\ref{fig:fig_sampling_scheme}\textbf{A}-\textbf{C} on the right hand side. Reevaluating requires almost no extra code functionality to perform; one need only to save a full precision trajectory, and run an unmodified molecular simulation code using the $u_1$ in order to calculate the new energies of stored snapshots. The analysis can be written in a line of code. We note that this method is even more general, in that the instantaneous work to change the potential energy function from $u_0$ to $u_1$ can be replaced by the non-reversible work $W$ to make the same change beginning from the same equilibrium conditions at either end state~\cite{jarzynski1997nonequilibrium,jarzynski1998equilibrium,crooks2000pathensemble}, allowing for non-equilibrium free energy calculations, an alternate approach. We do not detail non-equilibrium transformations here, and refer the reader to more advanced treatments~\cite{maragakis2008bayesian,oberhofer2005biased,procacci2015unbiased,shirts2003equilibrium,ytreberg2004singleensemble, gapsys2020large}, as our focus here is on equilibrium free energy techniques.

Although the Zwanzig equation is formally correct as long as the two states considered sample the same phase space volume, which is true for standard molecular models, it has some very important numerical issues that mean that it often performs badly for standard free energy calculations, even for small molecules~\cite{shirts2005comparison,lu2003appropriate}. One can show that if the standard deviation of the difference $\Delta u(\vec{q}) = u_1(\vec{q})-u_2(\vec{q})$ over all sampled $\vec{q}$ is large, which in this case, means only several times $k_BT$, then very few samples contribute to the average, and the answer will be both biased and extremely noisy~\cite{lelievre2010free}. Essentially, the method is dominated by contributions of rare snapshots~\cite{jarzynski2006rare, wu2005phasespace, wu2005phasespacea}.
